---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# plsmmLasso

<!-- badges: start -->

[![R-CMD-check](https://github.com/Sami-Leon/plsmmLasso/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Sami-Leon/plsmmLasso/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The objective of plsmmLasso is to facilitate the estimation of a Regularized Partial Linear Semiparametric Mixed-Effects Model (PLSMM) through a dictionary approach for modeling the nonparametric component. The model employs a set of bases functions and automatically selects them using a lasso penalty. Additionally, it conducts variable selection on the fixed-effects using another lasso penalty. The implementation also supports the inclusion of a random intercept.

## Installation

You can install the development version of plsmmLasso from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Sami-Leon/plsmmLasso")
```

## Example I: A Guide to Fitting, Visualization, and Inference

### Fitting the model

Here's a basic example using a simulated dataset to demonstrate how to utilize the main functions of the plsmmLasso package. This example assumes an effect of a grouping variable and different nonlinear functions for each group.

```{r example}
library(plsmmLasso)

# Simulate a dataset
set.seed(123)
data_sim <- simulate_group_inter(
  N = 50, n_mvnorm = 3, grouped = TRUE,
  timepoints = 3:5, nonpara_inter = TRUE,
  sample_from = seq(0, 52, 13), 
  cos = FALSE, A_vec = c(1, 1.5)
)
sim <- data_sim$sim

# Fit the data
x <- as.matrix(sim[, -1:-3])
y <- sim$y
series <- sim$series
t <- sim$t
bases <- create_bases(t)
lambda <- 0.0046
gamma <- 0.00001
plsmm_output <- plsmm_lasso(x, y, series, t,
  name_group_var = "group", bases$bases,
  gamma = gamma, lambda = lambda, timexgroup = TRUE,
  criterion = "BIC"
)
```

One of the most important output of the \code{plsmm_lasso()} function is the estimates of the fixed-effects.

```{r}
plsmm_output$lasso_output$theta
```

Here we see that some covariates have small values, but most are non-zeros. If we want more regularization for the fixed-effects we can use a larger value for lambda.

### Hyperparameter tuning

```{r}
lambda <- 0.1
plsmm_output <- plsmm_lasso(x, y, series, t,
  name_group_var = "group", bases$bases,
  gamma = gamma, lambda = lambda, timexgroup = TRUE,
  criterion = "BIC"
)
plsmm_output$lasso_output$theta
```

With a larger lasso penalty more coefficients are set to zero. The coefficients associated to the nonlinear functions are alpha.

```{r}
head(plsmm_output$lasso_output$alpha)
```

Similar behavior would be observe for the alphas if we were to increase the value of gamma.

To find optimal values for gamma and lambda we tune these hyperparameters using BIC-type criteria using the tune_plsmm function and a grid search.

```{r}
lambdas <- gammas <- round(exp(seq(log(1), log(1 * 0.00001),
              length.out = 5
)), digits = 5)

tuned_plsmm <- tune_plsmm(x, y, series, t,
                       name_group_var = "group", bases$bases,
                       gamma_vec = gammas, lambda_vec = lambdas, timexgroup = TRUE,
                       criterion = "BIC"
)
```

The function tuned_plsmm tries every possible combination of the values from lambdas and gammas and returns the model with the best BIC. This example is for illustration only, in practice a more exhaustive grid should be used.

### Plotting the results

The function plot_fit allows for the vizualisation of the estimated mean trajectories as well as the estimate of the nonlinear functions. By default only the observed time points are being used, to use predicted time points the argument predicted can be set to TRUE.

```{r}
plot_fit(x, y, series, t, name_group_var = "group", tuned_plsmm)

plot_fit(x, y, series, t, name_group_var = "group", tuned_plsmm, predicted = TRUE)
```

### Post-selection inference

To compute p-values on the fixed-effects the function debias_plsmm can be used.

```{r}
debias_plsmm(x, y, series, tuned_plsmm)
```

The function reports the original coefficients, debiased coeffcients, standard error, confidence intervals and p-values. These p-values are already ajudsted for the selection process of the lasso, and provide valid inference.

### Test on the nonlinear functions

Finally we can perform some tests on the nonlinear functions. The first element of the list is an overall test of equality, if the p-value is $<0.05$ we reject the null hypothesis of equality, and conclude that overall the two nonlinear functions are different. To obtain a comparison at each time point confidence bands are computed, and a figure displaying these confidence bands is generated. For the time points associated with confidence bands that include $0$ we can not reject the null hypothesis that the nonlinear functions are the same for this time point. The data.frame that is used to generate this figure can be found in the second element of the output list.

```{r}
test_f_results <- test_f(x, y, series, t,
 name_group_var = "group", tuned_plsmm,
 n_boot = 10
)
test_f_results[[1]]
head(test_f_results[[2]])
```

Similarly than for the plot_fit function the argument predicted can be changed to TRUE, to display the joint confidence bands as a continuous function of time rather than at the observed time points only.

```{r}
test_f_results <- test_f(x, y, series, t,
 name_group_var = "group", tuned_plsmm,
 n_boot = 10, predicted = TRUE
)
```

## Example II: Flexibility of the ...

### Model fitting without group-specific nonlinear functions

The model is flexible and if the nonlinear functions do not appear to be different the argument timexgroup can be set to FALSE. Here we simulate a data where the nonlinear functions are the same, and fit the data accordingly

```{r}
set.seed(123)
data_sim <- simulate_group_inter(
  N = 50, n_mvnorm = 3, grouped = TRUE,
  timepoints = 3:5, nonpara_inter = FALSE,
  sample_from = seq(0, 52, 13), 
  cos = FALSE, A_vec = c(1, 1.5)
)
sim <- data_sim$sim

# Fit the data
x <- as.matrix(sim[, -1:-3])
y <- sim$y
series <- sim$series
t <- sim$t
bases <- create_bases(t)
tuned_plsmm <- tune_plsmm(x, y, series, t,
                       name_group_var = "group", bases$bases,
                       gamma_vec = gammas, lambda_vec = lambdas, timexgroup = FALSE,
                       criterion = "BIC"
)

plot_fit(x, y, series, t, name_group_var = "group", tuned_plsmm)
```

As we can see now the estimates of the nonlinear functions are the same. It is also possible not to use any grouping variable for the name_group_var argument.

### Model fitting without grouping variable

Here we simulate a data that does not have an effect of a grouping variable (no difference in height of the overall mean trajectories) and the same nonlinear functions.

```{r}
set.seed(123)
data_sim <- simulate_group_inter(
  N = 50, n_mvnorm = 3, grouped = FALSE,
  timepoints = 3:5, nonpara_inter = FALSE,
  sample_from = seq(0, 52, 13), 
  cos = FALSE, A_vec = c(1, 1.5)
)
sim <- data_sim$sim

# Fit the data
x <- as.matrix(sim[, -1:-3])
y <- sim$y
series <- sim$series
t <- sim$t
bases <- create_bases(t)
tuned_plsmm <- tune_plsmm(x, y, series, t,
                       name_group_var = NULL, bases$bases,
                       gamma_vec = gammas, lambda_vec = lambdas, timexgroup = FALSE,
                       criterion = "BIC"
)

plot_fit(x, y, series, t, name_group_var = NULL, tuned_plsmm)
```

Only one figure is displyed now since there is no grouping variable.
